The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-04 paper/ms.tex
Move and expand observational discussion

An interesting, qualitative calculation can be made based on the velocity
variations seen in e.g. \citet{2019arXiv190209569F}. The velocity variations
in this work (admittedly found with respect to $J_{\phi}$ or radius) are
$\sim5\text{--}10\,\kms$. The vertical frequency of the thin-disk and
thick-disk orbits are $\sim 0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$,
respectively. By dimensional analysis, we therefore expect the midplane
offsets to be of order $\sim 57\text{--}170\,\pc$. We stress that this is a
rough calculation, but it shows that already we see evidence in the data from
velocities for midplane offsets on the order of what we see in both sets of
simulations.

@@

\subsection{Distribution Function Fitting}
\label{ssec:df}

>>>>

@@

\subsection{Estimate of Milky Way Midplane Offsets from Current Data}
\label{ssec:mw_data_midplane}

Systematic variations in $v_z$ were first noted as asymmetries in the local
velocity distribution towards the North and South Galactic Caps from the
radial velocity surveys of the Sloan Digital Sky Surveys
\citep{2012ApJ...750L..41W} and RAdial Velocity Experiment
\citep{2013MNRAS.436..101W}. Subsequently, \citet{2013ApJ...777L...5C} pointed
out suggestions of an oscillation in average vertical velocities of order
$5\,\kms$  on $\sim\,\kpc$ scales looking toward the Galactic anticenter. Work
by the {\em Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of these oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\text{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.

The vertical frequency of the thin-disk and thick-disk orbits are $\sim
0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$, respectively. By dimensional
analysis, and assuming a vertical velocity variation of $5\,\kms$, we
therefore expect the midplane offsets to be $\sim 57\textup{--}85\,\pc$. We
stress that this is a rough calculation, but it shows that already we see
evidence in the data from velocities for midplane offsets on the order of what
we saw in both sets of simulations.
